---
title: "Question 2"
author: "Gifty Osei"
date: "2024-11-01"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 2b.

In general, a nonlinear trend can look fairly linear when data contain noise. For example, the following code plots data generated from $y = f(x1, x2)$ = $x_1 ^2 + x_2^2 + \epsilon$, where $\epsilon \sim N(0,0.02^2)$.

```{r, warning=FALSE, message=FALSE}
set.seed(390)
n <- 100
# simulate 20 values from uniform(0,1)
x1 = runif(n,1,20)
x2= runif(n,1,20)

y = x1^2+x2*x2+rnorm(20,sd=0.02)
plot(y~x1, main = " Marginal plot of y vs x1", xlab = "x1", ylab = "y")
plot(y~x2,main = " Marginal plot of y vs x2", xlab = "x2", ylab = "y")



```
 We can see clearly form the marginal plots showing a linear trend but the overall model that generated the y values is nonlinear in x. We can see from the joint marginal plot that the are jointly nonlinear.
 
```{r, warning=FALSE, message=FALSE}
library(plotly)
plot_ly(x = ~x1, y = ~x2, z = ~y, type = "scatter3d", mode = "markers",
        marker = list(size = 3)) %>%
  layout(scene = list(xaxis = list(title = "x1"),
                      yaxis = list(title = "x2"),
                      zaxis = list(title = "y")))
```
 


## Question 2 

### a

```{r, warning=FALSE, message=FALSE}


# simulate data
n <- 100
x <- runif(n, 1, 10)  
beta_0 <- 3
beta_1 <- 2
sigma <- 3

# Generate y values based on the original model
epsilon <- rnorm(n, mean = 0, sd = sigma * x)
y <- beta_0 + beta_1 * x + epsilon

# transformations
y_prime <- y / x
x_prime <- 1 / x

# Check the variance of y_prime
var_y_prime <- var(y_prime)
print(sqrt(var_y_prime))

```
We can see that variance of $y^{'}$ is very close to the initial $\sigma$

### c

This is equivalent to OLS estimate because using the weights on the original data is like doing a $\frac{y}{x}$ transformation. We can see from the coeffiecients that there is not much change.

```{r, warning=FALSE, message=FALSE}
#  wls model
fit_wls <- lm(y ~ x, weights = 1 / x^2)

# OLS model
fit_ols_trans <- lm(y_prime ~ x_prime)

# Compare the coefficients
summary(fit_wls)$coefficients
summary(fit_ols_trans)$coefficients

```

## Question 3

```{r, warning=FALSE, message=FALSE}
library(faraway)
pander::pander(head(pipeline))

```


### a.
Fitting a linear model by regressing Lab on Field as;

```{r, warning=FALSE, message=FALSE}

fit_lm <- lm( Lab ~ Field, data = pipeline)



plot(fit_lm, which = 1)

```

We can see from the residual vs fitted plot that linearity assumption is checked but the residuals have a funnel like pattern. As x increases the variability increases.

### b

```{r, warning=FALSE, message=FALSE}

i = order(pipeline$Field)
npipe = pipeline[i,]
ff = gl(12,9)[-108]
meanfield = unlist(lapply(split(npipe$Field,ff),mean))
varlab = unlist(lapply(split(npipe$Lab,ff),var))
```


```{r, warning=FALSE, message=FALSE}

# Remove last point
meanfield <- meanfield[-length(meanfield)]
varlab <- varlab[-length(varlab)]

#  Log transform
log_meanfield <- log(meanfield)
log_varlab <- log(varlab)

#  model
var_model <- lm(log_varlab ~ log_meanfield)
summary(var_model)

# Extract coefficients for a0 and a1
## take exp because of the log 
a0 <- exp(coef(var_model)[1]) 

## slope - a1
a1 <- coef(var_model)[2]       

#   WLS as the inverse 
predicted_variance <- a0 * (pipeline$Field ^ a1)
weights <- 1 / predicted_variance

# Perform WLS regression of Lab on Field using the calculated weights
wls_model <- lm(Lab ~ Field, data = pipeline, weights = weights)
summary(wls_model)

```

```{r, warning=FALSE, message=FALSE}
# Assuming the pipeline dataset is available with columns Field and Lab
# Initial Scatter Plot
plot(pipeline$Field, pipeline$Lab, main = "Original Plot of Lab vs Field", 
     xlab = "Field", ylab = "Lab")

# 1. Square Root Transformations
# Square root of Lab vs. Field
sqrt_lab_model <- lm(sqrt(Lab) ~ Field, data = pipeline)
plot(pipeline$Field, sqrt(pipeline$Lab), main = "Square Root Transformation of Lab vs Field",
     xlab = "Field", ylab = "sqrt(Lab)")
abline(sqrt_lab_model, col = "blue")
# Diagnostic plots for sqrt(Lab) vs Field
par(mfrow = c(1, 2))
plot(sqrt_lab_model$fitted.values, sqrt_lab_model$residuals, main = "Residuals vs Fitted (sqrt(Lab))")
qqnorm(sqrt_lab_model$residuals, main = "QQ Plot of Residuals (sqrt(Lab))")
qqline(sqrt_lab_model$residuals)

# Square root of Field vs Lab
sqrt_field_model <- lm(Lab ~ sqrt(Field), data = pipeline)
plot(sqrt(pipeline$Field), pipeline$Lab, main = "Square Root Transformation of Field vs Lab",
     xlab = "sqrt(Field)", ylab = "Lab")
abline(sqrt_field_model, col = "blue")
# Diagnostic plots for Lab vs sqrt(Field)
par(mfrow = c(1, 2))
plot(sqrt_field_model$fitted.values, sqrt_field_model$residuals, main = "Residuals vs Fitted (sqrt(Field))")
qqnorm(sqrt_field_model$residuals, main = "QQ Plot of Residuals (sqrt(Field))")
qqline(sqrt_field_model$residuals)

# 2. Log Transformations
# Log of Lab vs Field
log_lab_model <- lm(log(Lab) ~ Field, data = pipeline)
plot(pipeline$Field, log(pipeline$Lab), main = "Log Transformation of Lab vs Field",
     xlab = "Field", ylab = "log(Lab)")
abline(log_lab_model, col = "blue")
# Diagnostic plots for log(Lab) vs Field
par(mfrow = c(1, 2))
plot(log_lab_model$fitted.values, log_lab_model$residuals, main = "Residuals vs Fitted (log(Lab))")
qqnorm(log_lab_model$residuals, main = "QQ Plot of Residuals (log(Lab))")
qqline(log_lab_model$residuals)

# Log of Field vs Lab
log_field_model <- lm(Lab ~ log(Field), data = pipeline)
plot(log(pipeline$Field), pipeline$Lab, main = "Log Transformation of Field vs Lab",
     xlab = "log(Field)", ylab = "Lab")
abline(log_field_model, col = "blue")
# Diagnostic plots for Lab vs log(Field)
par(mfrow = c(1, 2))
plot(log_field_model$fitted.values, log_field_model$residuals, main = "Residuals vs Fitted (log(Field))")
qqnorm(log_field_model$residuals, main = "QQ Plot of Residuals (log(Field))")
qqline(log_field_model$residuals)

# 3. Inverse Transformations
# Inverse of Lab vs Field
inv_lab_model <- lm(I(1/Lab) ~ Field, data = pipeline)
plot(pipeline$Field, 1/pipeline$Lab, main = "Inverse Transformation of Lab vs Field",
     xlab = "Field", ylab = "1/Lab")
abline(inv_lab_model, col = "blue")
# Diagnostic plots for 1/Lab vs Field
par(mfrow = c(1, 2))
plot(inv_lab_model$fitted.values, inv_lab_model$residuals, main = "Residuals vs Fitted (1/Lab)")
qqnorm(inv_lab_model$residuals, main = "QQ Plot of Residuals (1/Lab)")
qqline(inv_lab_model$residuals)

# Inverse of Field vs Lab
inv_field_model <- lm(Lab ~ I(1/Field), data = pipeline)
plot(1/pipeline$Field, pipeline$Lab, main = "Inverse Transformation of Field vs Lab",
     xlab = "1/Field", ylab = "Lab")
abline(inv_field_model, col = "blue")
# Diagnostic plots for Lab vs 1/Field
par(mfrow = c(1, 2))
plot(inv_field_model$fitted.values, inv_field_model$residuals, main = "Residuals vs Fitted (1/Field)")
qqnorm(inv_field_model$residuals, main = "QQ Plot of Residuals (1/Field)")
qqline(inv_field_model$residuals)

# Reset plotting layout
par(mfrow = c(1, 1))

```

